Tail associations and their impact on extinction risk

### 2019-11-03 MainText.Rmd
edits

Extinction risk is also very commonly estimated for age-structured models 
such as stochastic matrix models [@MorrisDoak2002; @Caswell2006] and 
integral projection models
[@Merowetal2014]. For such models, extinction risk depends, in part, on the long-term
population growth rate, a population parameter which in turn 
depends on the means, variances and covariances of population vital rates such 
as the survival and fecundity rates of age classes. Extinction risks may,
*a priori*, also depend on possible asymmetries of tail association between
population vital rates. For instance, it seems reasonable to imagine
a case in which both adult fecundity and juvenile survival rates 
depend on an environmental variable such as spring rainfall in similar ways, but 
only when that variable takes low values. When rainfall takes
typical or high values, fecundity and survival may be limited instead 
by different factors. This could produce left-tail dependence between these 
vital rates, possibly resulting in different extinction risk for the model
compared to a case in which vital rates were symmetrically associated in 
their tails but otherwise statistically similar.

@@

Because the deterministic, one-patch versions of our models showed undercompensatory
dynamics for $r<r_c$ (and therefore for $r$ on the left end of the range $(r_{min},r_{bf})$),
and showed overcompensatory dynamics for $r>r_c$, Figs \ref{fig_extrisk_allmodel_localdisp}
and \ref{SM-fig_extrisk_allmodel_globaldisp} suggest the hypothesis that the
opposite effects of tail association on extinction risk correspond to the domains
in which models exhibit under- versus overcompensatory dynamics. This was 
approximately, but not exactly, true. Careful inspection of Figs 
\ref{fig_risk_local_disp_verhulst_r_vary_scl_0.6} - \ref{fig_risk_local_disp_abrewer_r_vary_scl_0.6} and Figs \ref{fig_risk_global_disp_verhulst_r_vary_scl_0.6} - \ref{fig_risk_global_disp_abrewer_r_vary_scl_0.6}  
and the $r_c$ values of 
Table \ref{tab_model_summary} reveals that, for the Velhurst and Austin-Brewer models,
the switch occurs at a value of $r$ less than $r_c$ for these models. For the example
case of $d=0$, plots of extinction risk versus $r$ (Fig. \ref{fig_extrisk_d0}) seem to indicate that,
for the Ricker, Hassell, and Pennycuick models, left-tail association accentuates and
left-tail association mitigates extinction risk for undercompensatory dynamics ($r<r_c$),
whereas the reverse occurs for overcompensatory dynamics ($r>r_c$). Fig. \ref{fig_extrisk_d0}
reveals a similar reversal for the Maynard Smith, Velhurst and Austin-Brewer models; but 
it occurs at a value of $r$ close to but distinguishable from $r_c$.

@@

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between the values of an environmental variable in different
locations, $X_i(t)$ and $X_j(t)$, using standard methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used are recommended instead.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even through Pearson correlation
appears to inducate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends on both marginal information and association information.
It is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of  
variables, rather than reflecting true changes in association.
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018, @Valpine2014, @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, and how such 
associations may be changing.

>>>>

Extinction risk is also very commonly estimated for age-structured models 
such as stochastic matrix models [@MorrisDoak2002; @Caswell2000] and 
integral projection models
[@Merowetal2014]. For such models, extinction risk depends, in part, on the long-term
population growth rate, a population parameter which in turn 
depends on the means, variances and covariances of population vital rates such 
as the survival and fecundity rates of age classes. Extinction risks may,
*a priori*, also depend on possible asymmetries of tail association between
population vital rates. For instance, it seems reasonable to imagine
a case in which both adult fecundity and juvenile survival rates 
depend on an environmental variable such as spring rainfall in similar ways, but 
only when that variable takes low values. When rainfall takes
typical or high values, fecundity and survival may be limited instead 
by different factors. This could produce left-tail dependence between these 
vital rates, possibly resulting in different extinction risk for the model
compared to a case in which vital rates were symmetrically associated in 
their tails but otherwise statistically similar.

@@

Because the deterministic, one-patch versions of our models showed undercompensatory
dynamics for $r<r_c$ (and therefore for $r$ on the left end of the range $(r_{min},r_{bf})$),
and showed overcompensatory dynamics for $r>r_c$, Figs \ref{fig_extrisk_allmodel_localdisp}
and \ref{SM-fig_extrisk_allmodel_globaldisp} suggest the hypothesis that the
opposite effects of tail association on extinction risk correspond to the domains
in which models exhibit under- versus overcompensatory dynamics. This was 
approximately, but not exactly, true. Careful inspection of Figs 
\ref{SM-fig_risk_local_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} and Figs \ref{SM-fig_risk_global_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}  
and the $r_c$ values of 
Table \ref{tab_model_summary} reveals that, for the Velhurst and Austin-Brewer models,
the switch occurs at a value of $r$ less than $r_c$ for these models. For the example
case of $d=0$, plots of extinction risk versus $r$ (Fig. \ref{fig_extrisk_d0}) seem to indicate that,
for the Ricker, Hassell, and Pennycuick models, left-tail association accentuates and
left-tail association mitigates extinction risk for undercompensatory dynamics ($r<r_c$),
whereas the reverse occurs for overcompensatory dynamics ($r>r_c$). Fig. \ref{fig_extrisk_d0}
reveals a similar reversal for the Maynard Smith, Velhurst and Austin-Brewer models; but 
it occurs at a value of $r$ close to but distinguishable from $r_c$.

@@

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between the values of an environmental variable in different
locations, $X_i(t)$ and $X_j(t)$, using standard methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used are recommended instead.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even through Pearson correlation
appears to inducate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends on both marginal information and association information.
It is possible that recently observed changes in environmental
[e.g., @Black2018] and biological [e.g., @Koenig2016] synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of  
variables, rather than reflecting true changes in association.
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, and how such 
associations may be changing.
